Further Explorations of \suldiox~Aqueous Adsorption: Coordinated Bonding and Cyclic Hydrates

### 2011-12-01 Background.tex
Finishing up with edits returned from Geri. Getting ready to send out Draft 2.

A hydrated \suldiox~in an aqueous environment forms hydrogen bonds through the oxygens to nearby water-hydrogens, or interacts via the sulfur atom with water-oxygens. To further our analysis of the way in which \suldiox~coordinates its bonding to surface waters (those that lie in the topmost region of a gas/water interface), we adopt a naming scheme to denote the way in which the \suldiox~is hydrated by the surrounding waters. This naming scheme mimics a notational system developed during a previous study on water coordination by Buch et al,\cite{Buch 2005} and was subsequently used in more recent computational work.\cite{Walker2006b} In this naming system, a letter is used to designate the atom on a water molecule through which a hydrogen bond is formed to neighboring waters. Thus a bonding coordination of ``OOH'' designates two proton-acceptor bonding interactions through the water-oxygen, and a single proton-donor bonding interaction through a hydrogen. More recently, Baer et al. devised a nomenclature that explicitly enumerates the bonding to \suldiox~via the sulfur or oxygen atoms.\cite{Baer2010}

In this work we adopt a nomenclature scheme for \suldiox~in order to quantify hydrogen bonding through the acceptor \suldiox-oxygens, and the weaker bonding interactions from the \suldiox-sulfur to water-oxygens. Thus, an ``SOO'' coordinated \suldiox~molecule forms a single interaction through the sulfur atom to a neighboring water-oxygen, and two hydrogen bonds through either a single \suldiox-oxygen, or distributed with one hydrogen bond on each of the \suldiox-oxygens. Analysis of the distribution of \suldiox~coordinations will give insight to how \suldiox~binds to the water surface.

@@

Hydrated \suldiox~clusters have been studied extensively with many recent experiments and computations forming a clearer picture of \suldiox~bulk and surface behaviors.\cite{Baer2010, Tarbuck2005, Tarbuck2006, Ota2011, Bishenden1998, Hirabayashi2006, Steudel2009, Yang2002, Hayashi1985, Moin2011, Eckl2008, Jayne1990, Jayne1990a, Donaldson1995} At a water surface, it is now known that \suldiox~forms a complex with water during adsorption, and then subsequently absorbs into the interfacial region by reaction to form ionic sulfur species.\cite{Tarbuck2005, Tarbuck2006, Ota2011} The first attempt to elucidate the structure of surface hydrated \suldiox~by Baer et al. showed that the bonding coordination distributions of \suldiox~are altered relative to the bulk region, and that two coordinations dominate the distribution of bonding types: the ``SO'' and the ``SOO''. In the same work they then focused on the dominant coordination to determine the most likely cluster geometry using two and three-water clusters bound to the \suldiox. However, none of the studies have yet focused directly on the presence of an extended hydrating structure forming closed rings of molecular interactions involving the solvated \suldiox~molecules.

The cluster geometries found in this study imply a cyclic bonding structure through the two or three waters involved. ``Cyclic'' here is used to denote a closed loop formed by the intermolecular hydrogen bonds, S-O interactions, and covalent bonds of the molecules involved. Figure \ref{fig:cyclic-example} depicts one such cyclic structure showing the bonds beginning on the sulfur and returning through a \suldiox-oxygen.

@@

\subsection {Graph Theoretical Details}
The optimized geometry of the \suldiox~hydrates suggests cyclic bonding structures, but it remains a different story entirely when \suldiox~is placed in a dynamic environment such as in the course of MD simulations of an aqueous surface. Geometry optimization shows the formation of these cyclic hydrate structures with two or three waters in a static vacuum. Do the cyclic structures also form in the course of a dynamic bonding process on a simulated water surface, where extended hydrating structures influence \suldiox~and water behavior? To study the formation and behavior of cyclic hydrate structures we employ graph theoretical techniques on MD trajectory data. Previous use of graphs in molecular computations were applied to finding stable arrangements of water clusters, ice, hydrogen bonding, extracting topological molecular properties, and cyclic structure studies.\cite{Anick2002, Huber2007, Radhakrishnan1991, Shi2005, Garcia2004, McDonald1998}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/cycles/cycle-types-small.png}
		\caption{\suldiox~in various cyclic structures encountered during MD simulations with water. The cartoons show the five types of cyclic structures, numbered for reference. The cyclic structures may involve any number of waters, but here each structure is shown with three waters. Type \Rmnum{3} is formed by a \suldiox~with the ``SO'' bonding coordination, which is the most dominant bonding coordination encountered.}
		\label{fig:cyclic-structures}
	\end{center}
\end{figure}

>>>>

A hydrated \suldiox~in an aqueous environment forms hydrogen bonds through the oxygens to nearby water-hydrogens, or interacts via the sulfur atom with water-oxygens.\cite{Baer2010,Bishenden1998,Steudel2009} To further our analysis of the way in which \suldiox~coordinates its bonding to surface waters (those that lie in the topmost region of a gas/water interface), we adopt a naming scheme to denote the way in which the \suldiox~is hydrated by the surrounding waters. This naming scheme mimics a notational system developed during a previous study on water coordination by Buch et al,\cite{Buch 2005} and was subsequently used in more recent computational work.\cite{Walker2006b} In this naming system, a letter is used to designate the atom on a water molecule through which a hydrogen bond is formed to neighboring waters. Thus a bonding coordination of ``OOH'' designates two proton-acceptor bonding interactions through the water-oxygen, and a single proton-donor bonding interaction through a hydrogen. More recently, Baer et al. devised a nomenclature that explicitly enumerates the bonding to \suldiox~via the sulfur or oxygen atoms.\cite{Baer2010}

In this work we adopt a similar nomenclature scheme for \suldiox~in order to quantify hydrogen bonding through the acceptor \suldiox-oxygens, and the weaker bonding interactions from the \suldiox-sulfur to water-oxygens. Thus, an ``SOO'' coordinated \suldiox~molecule forms a single interaction through the sulfur atom to a neighboring water-oxygen, and two hydrogen bonds through either a single \suldiox-oxygen, or distributed with one hydrogen bond on each of the \suldiox-oxygens. Analysis of the distribution of these various \suldiox~coordinations will give insight to how \suldiox~binds to the water surface.

@@

Hydrated \suldiox~clusters have been studied extensively with several recent experiments and computations forming a clearer picture of \suldiox~bulk and surface behaviors.\cite{Baer2010, Tarbuck2005, Tarbuck2006, Ota2011, Bishenden1998, Hirabayashi2006, Steudel2009, Yang2002, Hayashi1985, Moin2011, Eckl2008, Jayne1990, Jayne1990a, Donaldson1995} At a water surface, it is now known that \suldiox~forms a complex with water during adsorption, and then subsequently absorbs into the interfacial region by reaction to form ionic sulfur species.\cite{Tarbuck2005, Tarbuck2006, Ota2011} The computational study to elucidate the structure of surface hydrated \suldiox~by Baer et al. showed that the bonding coordination distributions of \suldiox~at the surface are altered relative to the bulk region, and that two coordinations dominate the distribution of bonding types: the ``SO'' and the ``SOO''. In the same work they then focused on the dominant coordination to determine the most likely cluster geometry of di- and tri-hydrate species of \suldiox. However, that study, and others probing specific hydrate structures, were performed in gas phase, under idealized conditions following geometry optimizations. None of the studies have yet focused directly on the presence of an extended hydrating structure involving \suldiox~molecules at a water surface, forming closed rings of molecular interactions. Here we apply the ab initio molecular dynamics to recreate a microscopic water surface environment. By focusing on the occurence of a specific subset of \suldiox~bonding coordination types, we probe directly a certain hydrated \suldiox~in an aqueous interfacial environment.

The cluster geometries predicted in the study by Baer et al. imply a cyclic bonding structure through the two or three hydrating waters. ``Cyclic'' here is used to denote a closed loop formed by the intermolecular hydrogen bonds, S-O interactions, and covalent bonds of the molecules involved. Figure \ref{fig:cyclic-example} depicts one such cyclic structure showing the bonds beginning on the sulfur and returning through a \suldiox-oxygen.

@@

\subsection {Graph Theoretical Details}

As noted above, the optimized geometry of the \suldiox~hydrates suggests cyclic bonding structures. Geometry optimization shows the formation of these cyclic hydrate structures with two or three waters in the gas phase. A different story entirely has the potential to emerge when \suldiox~is placed in a dynamic environment such as in the course of MD simulations of an aqueous surface. Do the cyclic structures also form in the course of a dynamic bonding process on a simulated water surface, where extended hydrating structures influence \suldiox~and water behavior? To study the formation and behavior of cyclic hydrate structures we employ graph theoretical techniques on MD trajectory data. Previous use of graphs in molecular computations were applied to finding stable arrangements of water clusters, ice, hydrogen bonding, extracting topological molecular properties, and cyclic structure studies.\cite{Anick2002, Huber2007, Radhakrishnan1991, Shi2005, Garcia2004, McDonald1998}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/cycles/cycle-types-small.png}
		\caption{\suldiox~in various cyclic structures encountered during MD simulations with water. The cartoons show the five types of cyclic structures, numbered for reference. The cyclic structures may involve any number of waters, but here each structure is shown with three waters. Type \Rmnum{3} is formed by a \suldiox~with the ``SO'' bonding coordination, which is the most dominant bonding coordination encountered. Each hydrate structure is labeled with its corresponding bonding coordination type using the nomenclature described in the text.}
		\label{fig:cyclic-structures}
	\end{center}
\end{figure}

### 2011-12-01 Conclusions.tex
Finishing up with edits returned from Geri. Getting ready to send out Draft 2.

The adsorption of small gas molecules to water surfaces has been extensively studied over the past few decades. Much has been learned about the energies of hydrate configurations, and the kinetics of gaseous uptake into aqueous systems. Yet, the specific nature of the adsorption process, including the various geometries, hydrate species, and bonding pathways remains largely unknown. As a gas transitions into the liquid water phase, it passes through a fluid interfacial region that remains poorly understood. Our understanding of the processes and chemistry of the interface is still in its infancy, but we gain unique new insights that are key to understanding many environmentally important processes at aqueous surfaces.

>>>>

The adsorption of small gas molecules to water surfaces has been extensively studied over the past few decades. Much has been learned about the energies of hydrate configurations, and the kinetics of gaseous uptake into aqueous systems. Yet, the specific molecular nature of the adsorption process, including the various geometries, hydrate species, and bonding pathways remains largely unknown. As a gas transitions into the liquid water phase, it passes through a fluid interfacial region that remains poorly understood. Our understanding of the processes and chemistry of the interface is still in its infancy, but we are beginning to gain unique new insights that are key to understanding many environmentally important processes at aqueous surfaces.

### 2011-12-01 Coordination-Results.tex
Finishing up with edits returned from Geri. Getting ready to send out Draft 2.

Given the information we have about the frequencies of bonding coordination transitions, it is possible to draw a likely route of adsorption beginning with an unbound \suldiox. From the unbound coordination, the \suldiox~can bind to waters either through the sulfur or an oxygen to enter the ``S'' or ``O'' coordinations, respectively. At both temperatures we find that the coordination transition in Figure \ref{fig:coordination-transitions} from unbound to ``O'' occurs almost three times more than the transition from unbound to ``S''. We find two possibilities that may explain this difference. The ``O'' coordination may form more easily, but also break quickly after formation, accounting for the higher transition frequency. Otherwise, the ``O'' coordination may be the first step in adsorption of an unbound \suldiox, where a subsequent addition of an \suldiox-sulfur interaction to a water oxygen would lead to a transition to the most frequent coordination, ``SO''. In the latter case, any adsorption of \suldiox~may proceed through an oxygen binding, accounting for the increased unbound-``O'' transition frequency.

>>>>

Given the information we have about the frequencies of bonding coordination transitions, it is possible to draw a likely route of adsorption beginning with an unbound \suldiox. From the unbound coordination, the \suldiox~can bind to waters either through the sulfur or an oxygen to enter the ``S'' or ``O'' coordinations, respectively. At both temperatures we find that the coordination transition in Figure \ref{fig:coordination-transitions} from unbound to ``O'' occurs almost three times more than the transition from unbound to ``S''. We find two possibilities that may explain this difference. The single H-bonding of the ``O'' coordination may form more easily, but also break quickly after formation, accounting for the higher transition frequency. Otherwise, the ``O'' coordination may be the first step in adsorption of an unbound \suldiox, where a subsequent addition of an \suldiox-sulfur interaction to a water oxygen would lead to a transition to the most frequent coordination, ``SO''. In the latter case, any adsorption of \suldiox~proceeds through an oxygen binding, accounting for the increased unbound-``O'' transition frequency.

### 2011-12-01 Introduction.tex
Finishing up with edits returned from Geri. Getting ready to send out Draft 2.

One of Chemistry's many remaining ``black box'' events is the adsorption of gas molecules onto a water surface. Although gas uptake into aqueous systems occurs often environmentally and industrially, we still know very little about the process and the details of the adsorption reactions. How does gas initially bind to a water surface, and what steps are involved in the subsequent adsorption? How does an unbound gas molecule near a water surface affect the water to which it will bind? What is the structure of hydrating waters in the surface region, and how does a hydrated solute molecule behave differently than as a gas? Experiments to address these questions provide valuable information, but can never fully characterize microscopic events and behaviors. However, these systems can be fully characterized computationally, and when coupled to the previous experimental work can provide a much more complete picture of gaseous adsorption to aqueous surfaces.

As a model gas, \suldiox~has been used extensively in research because of its importance in commercial and environmental systems. \cite{Boniface2000,Jayne1990,Johns2011,Heber1997,Faloona2009,Clegg2001} Its molecular structure, high solubility in water, and relative abundance make it a pivotal compound in numerous aqueous atmospheric reactions. A complete picture of the \suldiox/\wat~adsorption process will aide in understanding gaseous adsorption on the many aqueous surfaces in the environment, as well as in understanding the fundamental nature of gases in water's surface region.

In this study we use ab initio quantum molecular dynamics (MD) techniques to model and simulate the hydrating structures that form around a surface-bound \suldiox~on water. We simulate a dynamic water surface with an extended hydrogen-bonding network that captures the variability of the \suldiox~hydrate structures, and the behavior of the water surface. The quantum MD technique described herein allows more accurate and realistic simulation than our previous classical MD.\cite{Shamay2011} It is also superior to small cluster DFT studies because it does not assume geometry optimized configurations, and the extended interactions of a water slab are incorporated. In our previous classical MD study we determined net orientational behavior of \suldiox~binding to a water surface, and the orientation of the waters as they respond to the presence of an adsorbing gas. Understanding the orientational behavior of molecules in the aqueous interfacial region during adsorption was a necessary first step to understanding the specific details of gas-binding and surface behavior.

@@

We believe this to be the first temperature study using quantum MD to study the binding of small gas molecules on a water surface. We show how temperature affects the bonding behavior of the surface-adsorbed \suldiox~to neighboring waters, and propose a sequential binding mechanism for \suldiox~adsorbing to a water surface. We also examine \suldiox~binding behavior when bound to the surface waters. Lastly, our analysis demonstrates an extended bonding structure of \suldiox~hydrates, as they are seen to preferentially form an extended cyclic ring structure through intermolecular bonds.

>>>>

The molecular nature of adsorption of gas molecules onto a water surface is one of the remaining largely uncharted territories of surface chemistry. Although gas uptake into aqueous systems occurs often environmentally and industrially, we still know very little about the process and the details of the adsorption reactions, and certainly less than what we know about gaseous adsorption on a solid surface. How does gas initially bind to a water surface, and what steps are involved in the subsequent adsorption? How does an unbound gas molecule near a water surface affect the water to which it will bind? What is the structure of hydrating waters in the surface region, and how does a hydrated solute molecule behave differently than as a gas? Experiments to address these questions provide valuable information, but have to date never fully characterize microscopic events and behaviors. However, these systems can be fully characterized computationally, and when coupled to the previous experimental work can provide a much more complete picture of gaseous adsorption to aqueous surfaces.

\suldiox~is a particularly important gas to use as a starting point and model system because of its importance in commercial and environmental systems. \cite{Boniface2000,Jayne1990,Johns2011,Heber1997,Faloona2009,Clegg2001} Its simple molecular structure, high solubility in water, and relative abundance make it a pivotal compound in numerous aqueous atmospheric reactions. A complete picture of the \suldiox/\wat~adsorption process will aide in understanding gaseous adsorption on the many aqueous surfaces in the environment, as well as in understanding the fundamental nature of gases in water's surface region.

In this study we use ab initio quantum molecular dynamics (MD) techniques to model and simulate the hydrating structures that form around a surface-bound \suldiox~on water. We simulate a dynamic water surface, complete with all the extended hydrogen-bonding interactions that capture the variability of the \suldiox~hydrate structures, and the behavior of the water surface molecules. The quantum MD technique described herein allows more accurate and realistic simulation than our previous classical MD.\cite{Shamay2011} It is also superior to small cluster DFT studies because it does not assume geometry optimized configurations, and the extended interactions of a water slab are incorporated. In our previous classical MD study we determined net orientational behavior of \suldiox~binding to a water surface, and the orientation of the waters as they respond to the presence of an adsorbing gas. Understanding the orientational behavior of molecules in the aqueous interfacial region during adsorption was a necessary first step to understanding the specific details of gas-binding and surface behavior.

@@

We believe this to be the first temperature study using quantum MD to study the binding of small gas molecules on a water surface. We show how temperature affects the bonding behavior of the surface-adsorbed \suldiox~to neighboring waters, and propose a sequential binding mechanism for \suldiox~adsorbing to a water surface. We also examine \suldiox~binding behavior when bound to the surface waters. Lastly, our analysis of a specific bonding arrangement demonstrates an extended bonding structure of \suldiox~hydrates, as they are seen to preferentially form an extended cyclic ring structure through intermolecular bonds.
